Testing for measurement invariance with respect to an ordinal variable

### 2013-04-04 ordinal/response.txt

Reviewer 1

Although I regard this work as rigorous, I have some doubts whether it adds sufficiently to the previous paper to warrant publication in its current form. I think it may need some additional extensions or additional work on power to render it publishable in Psychometrika. 
Such extensions may entail the following:

Comment 1.1. A comparison of the current approach with scenarios in which the factor analyst uses Modification Indices (MIs) to free parameters. In practice if one has, say, 5 age groups and a given parameter is found to lead to significant increases in model fit when restricted across groups, it is quite common to use MIs to detect the source(s) of the misfit and then to end up with, say one parameter for age groups 1-3  and another parameter for age groups 4 & 5 in a so-called partially invariant model (see Byrne et al., 1989). It would be interesting to see how that approach fares in comparison to the approaches described in the manuscript. I am also interested in reading the authors' ideas about the use MI in "ad hoc" model fitting.

Response 1.1. MIs are quite related to the proposed tests: the MI described by, e.g., Sorbom 1989 is equivalent to a Lagrange multiplier test, and the maximum Lagrange multiplier test statistic (computed across all potential thresholds of V) is within the family of statistics described by Merkle & Zeileis (2013).  However, as the reviewer states, MIs are typically applied only after a significant likelihood ratio test has been obtained (because the MI focuses on a single model parameter).  Our proposed statistics can be used across multiple parameters, potentially in place of the likelihood ratio test.  If the test is significant, one could use MIs or parameter fluctuation plots such as those in manuscript Figures 5 & 6.  We have added some brief detail on this relationship to the "Tests for Continuous V" section.

Comment 1.2. Given the widespread literature on ordinal factor analyses, it would be worthwhile to link the current modelling approach to ordinal factor analysis. Another major reason is that ordinal factor analysis naturally extends towards more standard IRT models. Therefore, by bridging the gap from the current modelling approach to ordinal factor analysis, the authors may also take steps to integrate it with the literature in measurement invariance in the IRT framework.

Response 1.2. We have added detail in the General Discussion on the extension of the proposed tests to other factor analysis models and to IRT.  In particular, our proposed tests can potentially extend to any psychometric model that assumes independent observations and that uses ML estimation.  Additionally, we cite some work on the application of similar tests to IRT.

Comment 1.3. Study of power under other violations of MI related to factor loadings or measurement intercepts.

Response 1.3. We now acknowledge that the obtained simulation results may not generalize to other measurement invariance violations or to other factor analysis models.  We feel that this suggestion would require multiple simulations that would result in a much longer paper, so we have not added new simulations to the manuscript.

Comment 1.4. Consider issues related to model comparison by means of AICs or BICs and the like. This is particularly relevant in light of the fact that monotonic violations of MI should be detectable in a more parsimonious manner than how it is often done in practice with likelihood ratio tests with many DFs.

Response 1.4. If we understand this comment correctly, the reviewer is suggesting that we compare the proposed statistics to the use of AIC or BIC (in addition to the LRT).  Under this framework, one would, e.g., reject the hypothesis of measurement invariance if the AIC for the "full" model were less than the AIC for the restricted model.  We have added the AIC strategy to our simulation, along with some other adjusted versions of the LRT (see Response 2.2).  The AIC turns out to have particularly low power.

Reviewer 2

Comment 2.1. On page 5 is the scaled sum of scores which is normalized by the square root of n and the covariance matrix of the scores I. This is fine theoretically, but what choice do you make in practice? In fact I think that here the problem is model misspecification, other than the biasing factor V. If the model used in ML is incorrect with respect to the number of factors for instance, then this will result in a inconsistent estimate of I. Why not use the sandwich by the late Halbert White (JASA, 1982 for instance)? This is easy to calculate and will improve the limiting distribution of the test statistics.  The choice of I is also missing in the simulations. Please be more specific what you did there.

Response 2.1. We have added an extended discussion of choices for the covariance matrix of scores, with references.  We have also clarified that, in the simulations, we used the information matrix.  Because our simulation models were correctly specified (except for the measurement invariance violation), we do not expect the choice of I to play a large role there.

Comment 2.2. The simulations show great improvement over the standard LR test on measurement invariance. And on page 10 the authors mention the results by Bentler and Bonett (1980) on LR tests picking up deviations too quickly. I think the alternative tests, adjustments of LR tests, like the one proposed in Yuan and Bentler (JASA, 1997). There the LR is downweighted ate rate O_p(1/n), which brings the results of the LR closer to the results of the authors' maxLM test. I would like to see a comparison with this or maybe some other adjusted LR test.

Response 2.2. We implemented three adjusted LRTs in our simulations: (i) the Yuan-Bentler 1997 adjustment suggested by the reviewer, (ii) the use of AIC (which, for nested models, essentially adjusts the LRT for complexity), and (iii) the Satorra-Bentler adjustment with correction for difference tests.  While these corrections downweight individual LRTs, the impact of the adjustments on difference tests is less clear.  Further, the Yuan-Bentler adjustment is seemingly designed for situations where the model is estimated via WLS, and there are no available corrections for difference tests.  We suspect that these issues are problematic, as our simulation results indicated that the Yuan-Bentler adjustment performs poorly in nearly all conditions.  Because the reason for this poor performance is not obvious (and because this issue is not focal for the paper), we only report results of statistics (ii) and (iii) in the manuscript.

Reviewer 3

Comment 3.1. The section entitled 'Theoretical Detail' (page 4) starts with the background for the score-based test statistics. However, from lines 18 onwards, there is a section on measurement invariance (starting with 'In studying measurement invariance, we are essentially testing...' up to line 19 on page 5, just before the subsection 'Tests for Continuous V'). I would suggest to put the section on measurement invariance testing in a separate (sub)section, and present this first, before providing the background for the test statistics.

Response 3.1.  We have moved these sections with thanks to the reviewer, so that measurement invariance comes before the theoretical discussion.

Comment 3.2. For readers unfamiliar with case-based scores, it may be useful to have a table with artificial values for some B(theta)_ij values; say for 20 subjects, and 4 parameters; in an additional column, there could be ordered continuous values for a V variable; in the next column, the V values could be binned to get an ordinal variable; referring to this table, it would be trivial to illustrate how the values of eg DM, CvM, maxLM, etc are computed...

Response 3.2. We agree that this would be useful for some readers, but we think that it would take up more space than it is worth.  For example, the calculation of ordinal test statistics would require both calculations within each bin and unique scaling (by (i/n)*(1 - i/n)) for each bin.  While these calculations are not difficult, they also consume space to write them out (especially if we included the non-ordinal statistics).  Additionally, besides the calculation of test statistics, it might be equally useful to show how the scores (s(theta; x_i)) can be used to calculate B(theta).  This implies that we would need both a table of scores (s(theta; x_i)) and a table of cumulative scores (B(theta)).  We feel that these ideas are represented more concisely and intuitively via their equations.

Comment 3.3. related to 2), I would suggest adding (perhaps in an appendix) some R code illustrating how (for example, for the values in the aforementioned table) the p-values can be computed for the various test statistics

Response 3.3. We have added some practical discussion of p-value computation to the "Choice of Test" subsection of the General Discussion.  We have also posted all the R code for replicating our results (p-values and otherwise) on our R-Forge page (listed in the "Computational Details" section).  We do not include code within the manuscript because the p-values associated with some of the statistics are computed via simulation (i.e., there is not a closed-form solution for critical values associated with the test statistic), and much of the underlying R code is already contained within R's strucchange package.

All minor points have been addressed, with thanks to the reviewer.

### 2013-04-04 ordinal/ordinal.Rnw
Revisions to ordinal paper: o Added new simulation results to ms o Addressed other reviewer comments in ms o Added draft of response letter o Updated references
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@194 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

The test statistics proposed here are derived from a family of tests
that were recently applied to the study of measurement invariance in
psychometric models \cite{KopZei10,MerZei12}.
In the following section, we provide an overview of the family and
describe the proposed statistics in detail.  Subsequently, we report on the
results of two simulation studies designed to compare the proposed
test statistics to existing tests of measurement
invariance.  Moreover, we illustrate the proposed statistics using
psychometric data on scales purported to measure youth gratitude.  
Finally, we provide some detail on the tests' use in practice.

@@

This section contains background on the theory underlying the
proposed statistics; for a more detailed
account, see \citeA{MerZei12}.

@@

In the section below, we review tests where
$V$ is continuous (and $\nu$ is unknown) before proceeding to the 
proposed tests for ordinal $V$.

@@

As mentioned previously, when $V$ is categorical with a
relatively-small number of categories, 
tests of measurement invariance typically proceed via multiple-group
models.  In this situation, we use likelihood ratio tests to compare a
model whose parameters differ across groups to a model whose
parameters are constrained to be equal across groups.  When $V$ is
continuous, however, multiple-group models usually cannot be used 
because there are no existing groups.  Instead, we can fit a model
whose parameters are restricted to be equal across all individuals and then
examine how individuals' scores ${\bm s}(\hat{{\bm \theta}}; x_i)$
fluctuate with their values of $V$.  If measurement invariance holds
with respect to $V$, then the scores should randomly fluctuate around
zero.  Conversely, if measurement invariance does not hold, then the
scores should systematically depart from zero.  These ideas are
related to those underlying the Lagrange multiplier test and are
discussed in detail by \citeA{MerZei12}. Here,
we review the tests' properties that are relevant for extending them
to the ordinal case.

To formalize the ideas discussed in the previous paragraph, we assume
that the observations are ordered 
w.r.t.\ $V$ (so that $v_i \le v_{i + 1}$) and define the $k$-dimensional
{\em cumulative score process} as 
\begin{equation} \label{eq:cumscore}
  {\bm B}(t; \hat {\bm \theta}) ~=~ \hat {\bm I}^{-1/2} n^{-1/2}
    \sum_{i = 1}^{\lfloor n t \rfloor} {\bm s}(\hat {\bm \theta}; x_i)
  \qquad (0 \le t \le 1)
\end{equation}
where $\lfloor nt \rfloor$ is the integer part of $nt$ and
$\hat I$ is some consistent estimate of the covariance matrix of the
scores, e.g., based on the information matrix or an outer product of the scores.
Equation~\eqref{eq:cumscore} simultaneously accounts for the
ordering of individuals w.r.t.\ $V$ and decorrelates the scores
associated with each of the $k$ model parameters (which allows us to
potentially make inferences separately for each individual model
parameter).  Using ideas similar to those that were outlined in the previous
paragraph, the cumulative score process associated with each
model parameter should randomly fluctuate around zero under
measurement invariance.  Further, there exists a functional central
limit theorem that allows us to make formal inference with this
cumulative score process.  Assuming that individuals are independent
and the usual ML regularity conditions hold, it is possible to show
that \cite{HjoKon02}
\begin{equation} \label{eq:fclt}
  {\bm B}(\cdot; \hat {\bm \theta}) ~\overset{d}{\rightarrow}~ {\bm B}^{0}(\cdot),
\end{equation}
where $\overset{d}{\rightarrow}$ denotes convergence in distribution
and ${\bm B}^{0}(\cdot)$ is a $k$-dimensional Brownian bridge.
Thus, we can construct tests of measurement invariance by comparing
the behavior of the cumulative score process to that of a Brownian bridge.
This is accomplished by comparing a scalar statistic associated with
the cumulative score process to the analogous statistic of a Brownian
bridge.

@@

%% Paragraph on test statistics
Specific test statistics commonly obtained under this framework
include the double maximum statistic
\begin{equation}
    \label{eq:dmax}
    \mathit{DM} = \max_{i = 1,\dots, n} \max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |,
\end{equation}
which essentially tests whether any component of the cumulative score
process strays too far from zero and is easily visualized.  This test
discards information 
related to multiple parameters fluctuating simultaneously, resulting
in it having relatively-low power for assessing measurement
invariance when multiple factor analysis parameters change
simultaneously \cite{MerZei12}.

Test statistics that exhibit better performance in such situations 
aggregate information across parameters and possibly also across
individuals.  These test statistics include
\begin{eqnarray}
    \label{eq:cvm}
    \mathit{CvM}     & = & n^{-1} \sum_{i = 1,\dots, n} \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2, \\
    \label{eq:maxlm}
    \max \mathit{LM} & = & \max_{i = \underline{i}, \dots, \overline{\imath}} ~
      \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
      \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2,
\end{eqnarray}
with the former being a Cram{\'e}r-von Mises statistic and the latter
corresponding to a ``maximum'' Lagrange multiplier test, where the
maximum is taken across all possible divisions of individuals into two
groups w.r.t.\ $V$.  Additionally, the $\max \mathit{LM}$ statistic is
scaled by the asymptotic variance $t (1 - t)$ of the process
${\bm B}(t, \hat {\bm \theta})$.  In simulations, \citeA{MerZei12} found 
that both tests perform well when assessing simultaneous changes
in multiple factor analysis parameters, with the
$\mathit{CvM}$ test being somewhat advantageous in their particular
simulation setup. These simulations included 
situations in which subsets of model parameters were tested; such
situations are handled 
by focusing only on those columns of ${\bm B}(\hat {\bm \theta})_{ij}$
that correspond to the parameters of interest.

@@

The theory described above was designed for situations where
$V$ is continuous, so that there is a unique ordering of individuals with respect to $V$.
However, in situations where $V$ is ordinal, there is only a partial ordering
of all individuals, i.e., observations with the same level of $V$ have no unique
ordering. (Note that the same also applies if $V$ is continuous in nature but is only
discretely measured leading to many ties.)

The ordinal statistics proposed here are
similar to those described in
Equations~\eqref{eq:dmax} and~\eqref{eq:maxlm}
above, except that we focus on ``bins'' of individuals at each level of the
ordinal variable.  That is, instead of aggregating over all $i = 1,
\dots, n$ individuals, we first compute cumulative proportions
$t_\ell$ ($\ell = 1, \dots, m-1$) associated with the first $m-1$
levels of $V$.  We then aggregate the 
cumulative scores only over $i_\ell = \lfloor n
\cdot t_\ell\rfloor$.  Test statistics related to~\eqref{eq:dmax}
and~\eqref{eq:maxlm} above can then be written as
\begin{eqnarray}
    \label{eq:wdm}
      \mathit{WDM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~ \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1/2}      
                             \max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |,\\
    \label{eq:maxlmo}
      \max \mathit{LM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~
                             \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
                             \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2.
\end{eqnarray}
Critical values associated with these test statistics can be obtained
by applying the same functionals to bins of a Brownian bridge, where the
bin sizes result in the cumulative proportions $t_\ell$ ($\ell=1, \dots,
m-1$) associated with the observed $V$.

@@

In this simulation, we demonstrate that the proposed test statistics are
sensitive to ordinal measurement invariance violations, moreso than
traditional statistics.  We generate data from a two-factor,
six-indicator model, with a measurement invariance violation occurring
in the unique variance parameters.  We
use the proposed test statistics to test for measurement invariance
simultaneously across the unique variances, which is similar to
comparing a tau-equivalent model to a congeneric model (e.g.,
\citeNP{VanLan00}).

We also compute two statistics that treat the ordinal
auxiliary variable as categorical.  These include the likelihood ratio
test of measurement 
invariance in the six unique variance parameters, along with the
unordered LM test from~\eqref{eq:lmuo}.

@@

Simulation results are presented in Figure~\ref{fig:sim1res}.  Rows of
the figure 
correspond to $n$, columns of the figure correspond to $m$, the x-axis
of each panel corresponds to $d$, and the y-axis of each panel
corresponds to power.  It is seen that one of the proposed test
statistics, the $\max \mathit{LM}_o$ statistic
from~\eqref{eq:maxlmo},
generally has the largest power to detect the ordinal measurement
invariance violations.  The other three tests are considerably closer
in power, with the second proposed ordinal statistic
(the double-max test from~\eqref{eq:wdm}) exhibiting the lowest power
at large violation magnitudes.  This is because the double-max
test discards information about multiple parameters changing together
at specific levels of the ordinal variable
(see \citeNP{MerZei12}, for related discussion), while the
three other tests under consideration make use of this information.
Finally, it is seen that, in the small $n$ and large $m$ conditions,
the likelihood ratio test exhibits large Type-I error rates (i.e.,
power greater than 0.05 at $d=0$).  This is because the likelihood
ratio test requires estimation of a multiple-groups model, which is very
unstable with large numbers of groups and small sample sizes as only
$n/m$ observations are available in each subsample.  The proposed
statistics are all of the LM-type and just require estimation of the single-group model,
leading to a clear advantage in these conditions.

\begin{figure}
\caption{Simulated power curves for the ordered and unordered
      $\max \mathit{LM}$ tests, the ordered double-max test, and the
      likelihood ratio test across three sample sizes $n$, three
      levels of the ordinal variable $m$, and measurement invariance
      violations of 0--1.5 standard errors (scaled by $\sqrt{n}$).}
\label{fig:sim1res}
\setkeys{Gin}{width=\textwidth}
<<sim1res, fig=TRUE, height=7, width=7>>=
trellis.par.set(theme = canonical.theme(color = FALSE))
mykey <- simpleKey(levels(sim1$test), points = TRUE, lines = TRUE)
print(xyplot(power ~ diff | nlevels + nobs, group = ~ test, data = sim1, 
             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
@
\end{figure}

@@

\subsection{Results}
Simulation results are presented in Figure~\ref{fig:sim2res}.  It is
observed that results are very consistent across the sample sizes
tested, implying that ``practical infinity'' is reached for this model
by $n=1200$.   We also observe a negative 
relationship between power and $m$; this is because the measurement
invariance violation occurred at only one level of the auxiliary
variable.  As $m$ increases (and $n$ is held
constant), the number of individuals violating measurement invariance
therefore decreases.  As a result, power to detect the violation decreases with
increasing $m$.

The more interesting result of this simulation lies in the comparison
of the four test statistics within each panel of the figure.  The
two ``unordered'' test statistics both have relatively-high power to
detect the measurement invariance violation, illustrating the result 
of \citeA{Ben80} and others that the
likelihood ratio test statistic picks out minor 
parameter discrepancies at large $n$.  In contrast, the two ordinal
test statistics that we proposed have considerably lower power, with
the $\mathit{WDM}_o$ statistic being the lowest and the $\max\mathit{LM}_o$
statistic being higher at larger values of $d$.  
This demonstrates that the proposed ordinal test
statistics can be especially useful at large sample sizes, where
traditional test statistics result in frequent significance.
Both statistics exhibited much lower power to detect a measurement
invariance violation that occurs only at a single level of $V$.

\begin{figure}
\caption{Simulated power curves for the ordered and unordered
      $\max \mathit{LM}$ tests, the ordered double-max test, and the
      likelihood ratio test across three sample sizes $n$, three
      levels of the ordinal variable $m$, and measurement invariance
      violations of 0--3 standard errors (scaled by $\sqrt{n}$)
      occurring at a single level (the $(1 +
      m/2)^{\text{th}}$ level) of the ordinal auxiliary variable.}
\label{fig:sim2res}
\setkeys{Gin}{width=\textwidth}
<<sim2res, fig=TRUE, height=7, width=7>>=
trellis.par.set(theme = canonical.theme(color = FALSE))
mykey <- simpleKey(levels(sim2$test), points = TRUE, lines = TRUE)
print(xyplot(power ~ diff | nlevels + nobs, group = ~ test, data = sim2, 
             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
@
\end{figure}

>>>>

The test statistics proposed here are derived from a family of tests
that were recently applied to the study of measurement invariance in
psychometric models \cite{KopZei10,MerZei13}.
In the following section, we provide an overview of the family and
describe the proposed statistics in detail.  Subsequently, we report on the
results of two simulation studies designed to compare the proposed
test statistics to existing tests of measurement
invariance.  Moreover, we illustrate the proposed statistics using
psychometric data on scales purported to measure youth gratitude.  
Finally, we provide some detail on the tests' use in practice.

\section{Measurement Invariance}

In the section below, we review the theory underlying tests where
$V$ is continuous (and $\nu$ is unknown).  We then propose novel tests
for ordinal $V$.

@@

This section contains background on the theory underlying the
proposed statistics; for a more detailed
account, see \citeA{MerZei13}.

@@

@@

As mentioned previously, when $V$ is categorical with a
relatively-small number of categories, 
tests of measurement invariance typically proceed via multiple-group
models.  In this situation, we use likelihood ratio tests to compare a
model whose parameters differ across groups to a model whose
parameters are constrained to be equal across groups.  When $V$ is
continuous, however, multiple-group models usually cannot be used 
because there are no existing groups.  Instead, we can fit a model
whose parameters are restricted to be equal across all individuals and then
examine how individuals' scores ${\bm s}(\hat{{\bm \theta}}; x_i)$
fluctuate with their values of $V$.  If measurement invariance holds
with respect to $V$, then the scores should randomly fluctuate around
zero.  Conversely, if measurement invariance does not hold, then the
scores should systematically depart from zero.  These ideas are
related to those underlying the Lagrange multiplier test and are
discussed in detail by \citeA{MerZei13}.  Additionally, these ideas
are related to those underlying Modification Indexes (e.g.,
\citeNP{Sor89}): the Modification Index is equivalent to a Lagrange
multiplier test, and the Lagrange multiplier test is contained in the
family described by \citeA{MerZei13}.  Here,
we focus on the tests' properties that are relevant for extending them
to the ordinal case.

To formalize the ideas discussed in the previous paragraph, we assume
that the observations are ordered 
w.r.t.\ $V$, with $x_{(i)}$ reflecting
the data for the individual who has the 
$i^{\text{th}}$-smallest value of $V$.  We then define the $k$-dimensional
{\em cumulative score process} as 
\begin{equation} \label{eq:cumscore}
  {\bm B}(t; \hat {\bm \theta}) ~=~ \hat {\bm I}^{-1/2} n^{-1/2}
    \sum_{i = 1}^{\lfloor n t \rfloor} {\bm s}(\hat {\bm \theta}; x_{(i)})
  \qquad (0 \le t \le 1)
\end{equation}
where $\lfloor nt \rfloor$ is the integer part of $nt$ and
$\hat I$ is some consistent estimate of the covariance matrix of the
scores.  Natural choices for $\hat I$ include the information matrix
(which we use in our applications and simulations below) or
alternatively some kind of outer product of the scores or sandwich
estimator to guard the inference against potential misspecification of
the model (see \citeNP{hub67} for the theoretical foundation and
\citeNP{zei06b} for a computational framework).  
Equation~\eqref{eq:cumscore} simultaneously accounts for the
ordering of individuals w.r.t.\ $V$ and decorrelates the scores
associated with each of the $k$ model parameters (which allows us to
potentially make inferences separately for each individual model
parameter).  Using ideas similar to those that were outlined in the previous
paragraph, the cumulative score process associated with each
model parameter should randomly fluctuate around zero under
measurement invariance.  Further, there exists a functional central
limit theorem that allows us to make formal inference with this
cumulative score process.  Assuming that individuals are independent
and the usual ML regularity conditions hold, it is possible to show
that \cite{HjoKon02}
\begin{equation} \label{eq:fclt}
  {\bm B}(\cdot; \hat {\bm \theta}) ~\overset{d}{\rightarrow}~ {\bm B}^{0}(\cdot),
\end{equation}
where $\overset{d}{\rightarrow}$ denotes convergence in distribution
and ${\bm B}^{0}(\cdot)$ is a $k$-dimensional Brownian bridge.
Thus, we can construct tests of measurement invariance by comparing
the behavior of the cumulative score process to that of a Brownian bridge.
This is accomplished by comparing a scalar statistic associated with
the cumulative score process to the analogous statistic of a Brownian
bridge.

@@

%% Paragraph on test statistics
Specific test statistics commonly obtained under this framework
include the double maximum statistic
\begin{equation}
    \label{eq:dmax}
    \mathit{DM} = \max_{i = 1,\dots, n} \max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |,
\end{equation}
which essentially tests whether any component of the cumulative score
process strays too far from zero and is easily visualized.  This test
discards information 
related to multiple parameters fluctuating simultaneously, resulting
in it having relatively-low power for assessing measurement
invariance when multiple factor analysis parameters change
simultaneously \cite{MerZei13}.

Test statistics that exhibit better performance in such situations 
aggregate information across parameters and possibly also across
individuals.  These test statistics include
\begin{eqnarray}
    \label{eq:cvm}
    \mathit{CvM}     & = & n^{-1} \sum_{i = 1,\dots, n} \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2, \\
    \label{eq:maxlm}
    \max \mathit{LM} & = & \max_{i = \underline{i}, \dots, \overline{\imath}} ~
      \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
      \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2,
\end{eqnarray}
with the former being a Cram{\'e}r-von Mises statistic and the latter
corresponding to a ``maximum'' Lagrange multiplier test, where the
maximum is taken across all possible divisions of individuals into two
groups w.r.t.\ $V$.  Additionally, the $\max \mathit{LM}$ statistic is
scaled by the asymptotic variance $t (1 - t)$ of the process
${\bm B}(t, \hat {\bm \theta})$.  In simulations, \citeA{MerZei13} found 
that both tests perform well when assessing simultaneous changes
in multiple factor analysis parameters, with the
$\mathit{CvM}$ test being somewhat advantageous in their particular
simulation setup. These simulations included 
situations in which subsets of model parameters were tested; such
situations are handled 
by focusing only on those columns of ${\bm B}(\hat {\bm \theta})_{ij}$
that correspond to the parameters of interest.

@@

The theory described above was designed for situations where
$V$ is continuous, so that there is a unique ordering of individuals with respect to $V$.
However, in situations where $V$ is ordinal, there is only a partial ordering
of all individuals, i.e., observations with the same level of $V$ have no unique
ordering. (Note that the same also applies if $V$ is continuous in nature but is only
discretely measured, leading to many ties.)

The ordinal statistics proposed here are
similar to those described in
Equations~\eqref{eq:dmax} and~\eqref{eq:maxlm}
above, except that we focus on ``bins'' of individuals at each level of the
ordinal variable.  That is, instead of aggregating over all $i = 1,
\dots, n$ individuals, we first compute cumulative proportions
$t_\ell$ ($\ell = 1, \dots, m-1$) associated with the first $m-1$
levels of $V$.  We then aggregate the 
cumulative scores only over $i_\ell = \lfloor n
\cdot t_\ell\rfloor$.  Test statistics related to~\eqref{eq:dmax}
and~\eqref{eq:maxlm} above can then be written as
\begin{eqnarray}
    \label{eq:wdm}
      \mathit{WDM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~ \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1/2}      
                             \max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |,\\
    \label{eq:maxlmo}
      \max \mathit{LM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~
                             \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
                             \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2,
\end{eqnarray}
resulting in a ``weighted'' double maximum statistic (weighted by the
asymptotic variance of the Brownian bridge) and an ordinal, maximum
Lagrange multiplier statistic.
Critical values associated with these test statistics can be obtained
by applying the same functionals to bins of a Brownian bridge, where the
bin sizes result in the cumulative proportions $t_\ell$ ($\ell=1, \dots,
m-1$) associated with the observed $V$.

@@

In this simulation, we demonstrate that the proposed test statistics are
sensitive to ordinal measurement invariance violations, moreso than
traditional statistics.  We generate data from a two-factor,
six-indicator model, with a measurement invariance violation occurring
in the unique variance parameters.  We
use the proposed test statistics to test for measurement invariance
simultaneously across the unique variances, which is similar to a test
of ``invariant uniquenesses'' (see \citeNP{VanLan00}). % Or comparing
                                        % parallel to tau-equivalent

The four ``traditional'' statistics that we consider generally treat
the ordinal auxiliary variable as categorical.  The first two are the
likelihood ratio test of measurement 
invariance in the six unique variance parameters and the
unordered LM test from~\eqref{eq:lmuo}.  Next, we consider the
Satorra-Bentler \citeyear{SatBen01} scaled likelihood ratio test
statistic with correction for difference testing.  Finally, we
consider the use of AIC \cite{Aka74} for detecting measurement
invariance.  In this context, we compute AIC for both the ``reduced'' 
model (all parameters constrained equal across groups) and the
``full'' model (unique variances allowed to differ between groups).
We then declare a measurement invariance violation if the AIC for the
full model is lower than the AIC for the reduced model.

@@

Simulation results comparing the ordinal tests to the unordered LM
test and the LRT are presented in Figure~\ref{fig:sim1res}.  Rows of
the figure 
correspond to $n$, columns of the figure correspond to $m$, the x-axis
of each panel corresponds to $d$, and the y-axis of each panel
corresponds to power.  It is seen that one of the proposed test
statistics, the $\max \mathit{LM}_o$ statistic
from~\eqref{eq:maxlmo},
generally has the largest power to detect the ordinal measurement
invariance violations.  The other three tests are considerably closer
in power, with the second proposed ordinal statistic
(the double-max test from~\eqref{eq:wdm}) exhibiting the lowest power
at large violation magnitudes.  This is because the double-max
test discards information about multiple parameters changing together
at specific levels of the ordinal variable
(see \citeNP{MerZei13}, for related discussion), while the
three other tests under consideration make use of this information.
Finally, it is seen that, in the small $n$ and large $m$ conditions,
the likelihood ratio test exhibits large Type-I error rates (i.e.,
power greater than 0.05 at $d=0$).  This is because the likelihood
ratio test requires estimation of a multiple-groups model, which is very
unstable with large numbers of groups and small sample sizes as only
$n/m$ observations are available in each subsample.  The proposed
statistics are all of the LM-type and just require estimation of the
single-group model, 
leading to a clear advantage in these conditions.

Simulation results comparing the ordinal tests to the Satorra-Bentler
scaled test statistic and to the AIC are presented in
Figure~\ref{fig:sim1bres}.  This figure has the same design as the
previous figure.  It is again seen that the $\max \mathit{LM}_o$
statistic exhibits the greatest power across the conditions.
Additionally, the Satorra-Bentler statistic has very large Type I
error rates at low-$n$/large-$m$ combinations but is comparable to the
ordinal double-max test at large $n$.  Finally, the ``power'' of the
AIC (measured by the proportion of times that the full model has a
lower AIC than the reduced model) is exceptionally low across all
conditions.  The full model's gain in fit is seldom worth the extra
parameters, at least according to AIC.

\begin{figure}
\caption{Simulated power curves for the ordered and unordered
      $\max\mathit{LM}$ tests, the ordered double-max test, and the
      likelihood ratio test across three sample sizes $n$, three
      levels of the ordinal variable $m$, and measurement invariance
      violations of 0--1.5 standard errors (scaled by $\sqrt{n}$).}
\label{fig:sim1res}
\setkeys{Gin}{width=\textwidth}
<<sim1res, fig=TRUE, height=7, width=7>>=
sim1.tmp <- sim1[sim1$test %in% c("maxLM_o","WDM_o","LRT","LM_uo"),]
sim1.tmp$test <- factor(sim1.tmp$test)
trellis.par.set(theme = canonical.theme(color = FALSE))
mykey <- simpleKey(levels(sim1.tmp$test), points = TRUE, lines = TRUE)
print(xyplot(power ~ diff | nlevels + nobs, group = ~ test, data = sim1.tmp, 
             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
@
\end{figure}

\begin{figure}
\caption{Simulated power curves for the ordered 
      $\max\mathit{LM}$ tests, the ordered double-max test, the
      Satorra-Bentler (2001) scaled likelihood ratio
      test, and the use of AIC 
      across three sample sizes $n$, three 
      levels of the ordinal variable $m$, and measurement invariance
      violations of 0--1.5 standard errors (scaled by $\sqrt{n}$).}
\label{fig:sim1bres}
\setkeys{Gin}{width=\textwidth}
<<sim1bres, fig=TRUE, height=7, width=7>>=
sim1.tmp <- sim1[sim1$test %in% c("maxLM_o","WDM_o","LRT_sb","AIC"),]
sim1.tmp$test <- factor(sim1.tmp$test)
trellis.par.set(theme = canonical.theme(color = FALSE))
mykey <- simpleKey(levels(sim1.tmp$test), points = TRUE, lines = TRUE)
print(xyplot(power ~ diff | nlevels + nobs, group = ~ test, data = sim1.tmp, 
             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
@
\end{figure}

@@

\subsection{Results}
Simulation results for the two ordinal test statistics, the unordered LM
test, and the LRT are presented in Figure~\ref{fig:sim2res}.  It is
observed that results are very consistent across the sample sizes
tested, implying that ``practical infinity'' is reached for this model
by $n=1200$.   We also observe a negative 
relationship between power and $m$; this is because the measurement
invariance violation occurred at only one level of the auxiliary
variable.  As $m$ increases (and $n$ is held
constant), the number of individuals violating measurement invariance
therefore decreases.  As a result, power to detect the violation decreases with
increasing $m$.

The more interesting result of Figure~\ref{fig:sim2res} lies in the comparison
of the four test statistics within each panel of the figure.  The
two ``unordered'' test statistics both have relatively-high power to
detect the measurement invariance violation, illustrating the result 
of \citeA{Ben80} and others that the
likelihood ratio test statistic picks out minor 
parameter discrepancies at large $n$.  In contrast, the two ordinal
test statistics that we proposed have considerably lower power, with
the $\mathit{WDM}_o$ statistic being the lowest and the $\max\mathit{LM}_o$
statistic being higher at larger values of $d$.

Simulation results for the two ordinal test statistics, the Satorra-Bentler
scaled test statistic, and the AIC are presented in
Figure~\ref{fig:sim2bres}.  It is observed that the Satorra-Bentler
statistic exhibits performance that is similar to the usual LRT,
with the two ordinal statistics having lower power.  The AIC, which
generally exhibited low power even for Simulation 1, continues to
exhibit low power here.  However, if $n$ or $d$ is large enough, the
power exhibited by AIC is larger than the power of the ordinal test
statistics.

These results demonstrate that the proposed ordinal test
statistics can be especially useful at large sample sizes, where
traditional test statistics result in frequent significance.
Both statistics exhibited much lower power to detect a measurement
invariance violation that occurs only at a single level of $V$.
